---
title: "ExploratoryVisualization"
author: "jhs"
date: "2025-12-09"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

We will begin exploring our county data to understand the distributions and tendencies. To do so, we must set up our workspace with relevant packages and import our processed data.

```{r setup, echo=TRUE, include=FALSE}
#install.packages(tidyverse);install.packages(here);install.packages(ggthemes)
library(tidyverse)
library(here)
library(ggthemes)
here()
NCData <- read.csv(here('Data','Processed','NCDataClean.csv'),stringsAsFactors = T)

my_theme <- theme_calc()  + 
  theme(
    #line =              element_line(),
    #rect =              element_rect(),
    text =              element_text(family="mono"),
    
    # Modified inheritance structure of text element
    #plot.title =        element_text(),
    #axis.title.x =      element_blank(),
    #axis.title.y =      element_blank(),
    #axis.text =         element_text(),
    
    # Modified inheritance structure of line element
    #axis.ticks =        element_line(),
    #panel.grid.major =  element_line(),
    #panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    #plot.background =   element_rect(color =),
    #panel.background =  element_rect(),
    #legend.key =        element_rect(),
    
    # Modifiying legend.position
    #legend.position = 'right',
    
    #complete = TRUE
    )

```

With our data loaded, we can begin to explore the distributions and tendencies of our data.

Calling the summary function on the data reveals the basic dimensions of the data: minimum, quartiles, median, mean and maximum.

```{r}
summary(NCData)
```

We can see right off the bat that the difference in population between the smallest and largest count is about a factor of 1,000. We can assume the distribution of counts of population is right skewed because the mean is larger than the median. The graph, figure 1 below, proves this visually.

```{r}
PopPlot <- ggplot(data = NCData)+
  geom_histogram(aes(x= Total.Population), show.legend = T,
                 binwidth = 10000)+
  labs(title = "Figure 1: Total Population Distribution",
       subtitle = "Right Skew Confirmed; Bin Width = 10,000",
       x = "Total Population",
       y = "Number of Counties") 
PopPlot
```

Household incomes do not demonstrate the same magnitude in range: the minimum average household income of the counties is about a third of the maximum average household income. In the case of median household incomes, the difference is much smaller.

```{r}

```

Walkability Index Scores and percentages of population in different designations of walkability imply that walkability across the state is low; it appears the majority of counties have half their population residing in "least walkable" and "below average walkable" conditions.

Lastly, we see that approximately three quarters of counties in North Carolina have below 10% of their population living within Flood Hazard Areas. However, at least one country is living with approximately 94.5% of its population in Flood Hazard Areas.

Let's explore these variables visually.
